A general and efficient representation of Ancestral Recombination Graphs

### 2022-06-16 paper.tex
Minor edits improving ag dataset part

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype data. For example,
the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in pigs is also less than 100 with each female and male
routinely producing respectively more than 10 and 5000 progeny each \citep{Hall2016,Porcnic2016}.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

>>>>

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype data \citep[e.g.][]{Cesarani2022}. For example,
the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is the bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.
